---
title: "Regular Expressions Tech Talk"
output: html_notebook
---

### Introduction

_Regular expressions are a notation for describing sets of character strings_

#### Possible reasons people don’t learn regex

  - They don’t realize all of the places they could be using regular expressions until they actually know how to use them
  - They have seen something that looks like this: http://www.ex-parrot.com/~pdw/Mail-RFC822-Address.html

#### What We'll Cover

  - Background
  - Regex basics
  - Modifying multiple similar (but different) parts of a string
  - Extracting something from a string based on the characters around it
  - Multiple places you can use regex

### Background

#### Regex Engines

  - POSIX Basic Regular Expressions ([BRE](https://en.wikipedia.org/wiki/Regular_expression#POSIX_basic_and_extended))
  - POSIX Extended Regular Expressions ([ERE](https://en.wikipedia.org/wiki/Regular_expression#POSIX_basic_and_extended))
    - This is what R uses by default
  - Perl Compatible Regular Expressions ([PCRE](https://en.wikipedia.org/wiki/Perl_Compatible_Regular_Expressions))
    - R has the ability to use this and it is encouraged
    - Python, Java, Javascript
    
We're going to focus on Perl because it is pretty similar for most things.


### Regex Basics

We'll setup a function to test this stuff out easily
```{r}

findAll <- function(pattern, x) {
  
  regmatches(x = x, 
             m = gregexpr(pattern = pattern, 
                          text = x, 
                          perl = T))[[1]]
  
}

```

To start, we're going to be matching text from the following string
```{r}
x <- 'the cat walked down the slippery catwalk'
```

We can very simply find the word 'cat' twice
```{r}
findAll(pattern = 'cat', x = x)
```

#### Special Characters

##### Dot
We can match 'cat' followed by any character
```{r}
findAll(pattern = 'cat.', x = x)
```

##### Brackets
We can match a character from a set of characters
```{r}
findAll(pattern = '[abc]', x = x)
```

Combining what we have seen so far
```{r}
findAll(pattern = '[abc].', x = x)
```

Within the brackets, you can also define ranges
```{r}
findAll(pattern = '[a-z]', x = x)
```

##### Caret
We can negate characters in square brackets
```{r}
findAll(pattern = '[^abc]', x = x)
```

##### Caret (Again) and Dollar Sign
Outside of square brackets, we can indicate the beginning of a string
```{r}
findAll(pattern = '^the ...', x = x)
```

Note that if we took out the caret, we would have two matches

Similarly, we can indicate the end of the string
```{r}
findAll(pattern = '...walk$', x = x)
```

##### Asterisk and Plus Sign
We can match either 0 or more of the previous character
```{r}
findAll(pattern = 'slip*ery', x = x)
```

Similarly, we can match either 1 or more of the previous character
```{r}
findAll(pattern = 'slip+ery', x = x)
```

Note that this can be useful when you have an arbitrary number of repeated characters

For example, the following will match the whole string
```{r}
findAll(pattern = '.*', x = x)
```

##### Curly Bracket
We can match between m and n more of the previous character
```{r}
findAll(pattern = '.{1,3}', x = x)
```

##### Shorthand Characters
You can use the following instead of using characters directly
```
\\w Word characters
\\W Non-word characters
\\d Digit characters
\\D Non-digit characters
\\s Whitespace characters
\\S Non-whitespace characters
\\v Vertical whitespace characters
\\V Non-vertical whitespace characters
\\h Horizontal whitespace characters
\\H Non-horizontal whitespace characters
```

##### Parentheses and Pipe
We can match with 'OR' conditions
```{r}
findAll(pattern = 'the (cat|slippery)', x = x)
```

With parentheses, can run subexpressions and capture their output
NOTE: R does not do a great job of this
One option using base R
```{r}
gsub(pattern = 'the ([a-z]+)', 
     replacement = 'the \\U\\1', # \\U for uppercase \\1 for matching the first group
     x = x, 
     perl = T)
```

##### Lookbehind
We can match one pattern and then look for additional matches before
Note that we no longer need the 'the ' in the replacement
```{r}
gsub(pattern = '(?<=the )([a-z]+)', 
     replacement = '\\U\\1', # \\U for uppercase \\1 for matching the first group
     x = x, 
     perl = T)
```

##### Lookahead
We can match one pattern and then look for additional matches after
```{r}
gsub(pattern = '([a-z]+ )(?=cat\\s)', 
     replacement = '\\U\\1', # \\U for uppercase \\1 for matching the first group
     x = x, 
     perl = T)
```

#### Practice
To practice the basics, check out https://regexone.com/, which has a bunch of interactive examples.

#### Where to use Regex

You can use regex in a number of places

- [R](https://stat.ethz.ch/R-manual/R-devel/library/base/html/regex.html)
- [Python](https://docs.python.org/2/library/re.html)
- [Sublime](http://docs.sublimetext.info/en/latest/search_and_replace/search_and_replace_overview.html#using-regular-expressions-in-sublime-text)
- [Rstudio](https://www.r-bloggers.com/regular-expressions-in-r-vs-rstudio/)
- [vim](http://vimregex.com/), [sed](https://www.tutorialspoint.com/unix/unix-regular-expressions.htm)
- [Vimium](http://vimium.github.io/)

##### Tips with Vim

```
:range s[ubstitute]/pattern/string/cgiI
```

Add a range at the beginning

| Specifier | Description |
| -- |:-- |
| `number` | an absolute line number |
| `.` | the current line |
| `$` | the last line in the file |
| `%` | the whole file. The same as 1,$ |
| `'t` | position of mark "t" |
| `/pattern[/]` | the next line where text "pattern" matches. |
| `?pattern[?]` | the previous line where text "pattern" matches |
| `\/` | the next line where the previously used search pattern matches |
| `\?` | the previous line where the previously used search pattern matches |
| `\&` | the next line where the previously used substitute pattern matches |

Add one of the following options at the end

`c` - Confirm each substitution

`g` - Replace all occurrences in the line (without g - only first).

`i` - Ignore case for the pattern.

`I` - Don't ignore case for the pattern.

So when you want to replace something
```
:%s/WORD_TO_MATCH/REPLACEMENT/g
```
